`traveltime`: an R package to calculate travel time across a landscape from user-specified locations

### 2025-05-02 paper/paper.qmd
update existing material into verixiv strucutre

---
title: "`traveltime`: an R package to calculate travel time across a landscape from user-specified locations"
format:
  # md: default
  # wordcount-html:
  # html:
  #   keep-md: true
  #   fig-height: 4
  #   fig-align: center
  #   fig-format: png
  #   dpi: 300
  #pdf:
  docx:
    keep-md: true
    fig-height: 4
    fig-align: center
    dpi: 300
    fig-format: png
# header-includes:
#   \usepackage{lineno} \linenumbers
#   \usepackage{hanging}
tags: 
  - R
  - geographic information systems
  - spatial analysis
author:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 4"
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
 - name: University of Western Australia, WA, Australia
   index: 4
date: 2025-04-07
bibliography: paper.bib
---

@@

# Summary

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations (geographic coordinates). The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations. We envisage this software having diverse applications including: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].

@@

`traveltime` provides a spatial interface using object classes from the `terra` package [@terra]. Travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017]. To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` internally uses `malariaAtlas` [@pfeffer2018malariaatlas] to download surfaces for the area of interest; though users can also supply any other friction surface.

`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and documented at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package are accompanied by citing the underlying methodological work as well [@weiss2018global; @weiss2020global].

# Context

Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care [@hulland2019travel; @weiss2020global] have generated much interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. Here we extend that work to enable travel time calculations from any arbitrary set of locations and friction surface.

# Example: walking from public transport in Singapore

Here we provide a very brief example to calculate and map the walking travel time from rail transport stations across Singapore. [Complete code for this example is available as a vignettte in package documentation](https://idem-lab.github.io/traveltime/articles/singapore.html).

## Prepare data and friction surface

@@

We download `singapore_shapefile`, a polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata] as our area of interest. Our points are the `traveltime::stations` data, containing coordinates of all LRT and MRT station exits in Singapore [@singdata]. We then download a walking travel friction surface for Singapore:

```{r}
#| echo: false
#| output: false
#| label: get-basemap
library(terra)
library(geodata)

@@

```{r}
#| echo: false
#| output: false
#| label: show-stations
library(traveltime)
head(stations)
```

```{r}
#| label: download-friction-surface
#| echo: true
#| output: false
friction_singapore <- traveltime::get_friction_surface(
    surface = "walk2020",
    extent = singapore_shapefile
  )|> 
  terra::mask(singapore_shapefile)
```

(Alternatively, we could use `surface = "motor2020"` for motorised travel. We're also only interested in walking *on land* so we `mask` out areas outside of the land boundary in `singapore_shapefile`)

```{r }
#| echo: false
#| output: false
#| label: show-friction-surface-raster
friction_singapore
```

@@

## Calculate and plot the travel time

@@

# Opportunities for future development

The `traveltime` package is immediately suitable to a range of application. Nonetheless, we see opportunities to build the package utility through:

@@

Firstly, `traveltime` currently facilitates access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2]. Although the user can presently supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global]. As landscapes are dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell. We underline however that a user can provide their own higher resolution friction surface at present if desired.

[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator

>>>>

---
title: "`traveltime`: an R package to calculate travel time across a landscape from user-specified locations"
format:
  # md: default
  # wordcount-html:
  # html:
  #   keep-md: true
  #   fig-height: 4
  #   fig-align: center
  #   fig-format: png
  #   dpi: 300
  #pdf:
  docx:
    keep-md: true
    fig-height: 4
    fig-align: center
    dpi: 300
    fig-format: png
# header-includes:
#   \usepackage{lineno} \linenumbers
#   \usepackage{hanging}
tags: 
  - R
  - geographic information systems
  - spatial analysis
author:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 4"
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
 - name: University of Western Australia, WA, Australia
   index: 4
date: 2025-05-02
bibliography: paper.bib
---

@@

# Abstract

Understanding and mapping the time to travel among locations is useful for many activities from urban planning to public health and myriad others. Here we present a software package --- `traveltime` --- written in and for the language R. `traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations defined by geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations. We envisage this software having diverse applications including: estimating sampling bias, allocating defibrillators, setting health districts, or mapping access to vehicle chargers and agricultural facilities. The work-flow requires two key steps: preparing a friction surface for the area of interest, and then calculating travel time over that surface for the points of interest. `traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and documented at <https://idem-lab.github.io/traveltime/>.

# Introduction

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care [@hulland2019travel; @weiss2020global] have generated much interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. Here we extend that work to enable travel time calculations from any arbitrary set of locations and friction surface.

We present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations defined by geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations.

# Methods

## Implementation

`traveltime` is an R [@Rref] package and requires installation of R version 4.1 or a more recent version `traveltime` provides a spatial interface using object classes from the `terra` package [@terra]. Travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017]. To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` internally uses the R package `malariaAtlas` [@pfeffer2018malariaatlas] to download surfaces for the area of interest; though users can also supply any other friction surface raster.

## Operation

@@

### Installation

`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and documented at <https://idem-lab.github.io/traveltime/>. It can be installed in R as follows:

``` r
install.packages("traveltime", repos = c("https://idem-lab.r-universe.dev"))
```

### Example: walking from public transport in Singapore

Here we provide an example to calculate and map the walking travel time from rail transport stations across Singapore. [Complete code for this example is available as a vignettte in package documentation](https://idem-lab.github.io/traveltime/articles/singapore.html).

#### Prepare data and friction surface

@@

We download `singapore_shapefile`, a polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata] as our area of interest:

```{r}
#| echo: true
#| output: true
#| label: get-basemap
library(terra)
library(geodata)

@@

Next we use the function `traveltime::get_friction_surface` to retrieve a walking friction surface for our area of interest. Alternatively, we could use `surface = "motor2020"` for motorised travel. We're also only interested in walking *on land* so we `mask` out areas outside of the land boundary in `singapore_shapefile`. Users supplying their own friction surfaces do not need to download one in this fashion, only ensure that it is in `SpatRaster` format.

```{r}
#| label: download-friction-surface
#| echo: true
#| output: true
friction_singapore <- traveltime::get_friction_surface(
    surface = "walk2020",
    extent = singapore_shapefile
  )|> 
  terra::mask(singapore_shapefile)

friction_singapore
```

Our points are the `traveltime::stations` data, containing coordinates of all LRT and MRT station exits in Singapore [@singdata]:

```{r}
#| echo: true
#| output: true
#| label: show-stations
library(traveltime)
head(stations)
```

@@

#### Calculate and plot the travel time

@@

# Discussion

The `traveltime` package is immediately suitable to be used 'out-of-the-box' with any set of coordinates, in any part of the globe. We envisage this software having diverse applications including: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting]. Nonetheless, we see opportunities to build the package utility through:

@@

Firstly, `traveltime` currently facilitates access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator). Although the user can presently supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global]. As landscapes are dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell. We underline however that a user can provide their own higher resolution friction surface at present if desired.

At the other end of the scale, as the area of interest increases, the size of the matrix of calculations necessary increases exponentially, making significant memory demands for analyses over large landscapes (e.g. analyses over Africa required \~ 72 GB RAM to run successfully). Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.

Lastly, although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package are accompanied by citing the underlying methodological work as well [@weiss2018global; @weiss2020global].
